---
title: "Cache and exported copies"
---

```{r, include = FALSE}
# Nothing on this page needs to run, and running it would download data
# during the site build, so every chunk is display-only.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The first time you ask for a municipality, `cnefetools` downloads it from
IBGE and keeps a copy on your disk, so later calls reuse that copy instead
of downloading it again, in the same R session or in any other. This page
explains where that copy lives, how to move it somewhere else, how to clear
it, and how to keep a permanent copy of your own with `cnefe_export()`.

## Where the cache lives

The cache folder is chosen from three places, and the first one that is set
wins:

| Order | Where it comes from | What it affects |
|:-----:|:--------------------|:----------------|
| 1 | The `cache_dir` argument of the call | That call only |
| 2 | The `CNEFETOOLS_CACHE_DIR` environment variable | Every call, in every session that sees the variable |
| 3 | `tools::R_user_dir("cnefetools", "cache")` | Everything, when neither of the above is set |

If you've never set either of the first two, your cache is in the default
folder, and you can see where that is with:

```{r}
tools::R_user_dir("cnefetools", which = "cache")
```

Inside that folder there's one subfolder per CNEFE edition, and in it one
gzipped CSV per municipality you've used. The census tract files that
`tracts_to_h3()` and `tracts_to_polygon()` need are kept in `sc_assets/`, one
Parquet file per state:

```
cnefetools/
└── 2022/
    ├── 2919207_LAURO_DE_FREITAS.csv.gz
    ├── 3550308_SAO_PAULO.csv.gz
    └── sc_assets/
        ├── sc_29.parquet
        └── sc_35.parquet
```

A large municipality takes up a fair amount of space (São Paulo is about 177
MB), so if you work with many municipalities it's worth thinking about which
disk the cache should be on.

## Moving the cache to another disk

If you want the cache somewhere else for good, set the environment variable in
your `.Renviron` file. You can open it with `usethis::edit_r_environ()`, add a
line like the one below, save the file and restart R:

```
CNEFETOOLS_CACHE_DIR=D:/cnefe_cache
```

From then on every function uses that folder without you having to pass
anything.

If you only want to redirect one call, for example to try something on an
external drive without changing your setup, pass `cache_dir` instead:

```{r}
library(cnefetools)

counts <- cnefe_counts(2919207, cache_dir = "E:/cnefe_cache")
```

Every function that reads CNEFE data accepts `cache_dir`: `read_cnefe()`,
`cnefe_counts()`, `compute_lumi()`, `tracts_to_h3()`, `tracts_to_polygon()` and
`cnefe_export()`.

Changing the folder doesn't move what's already in the old one. The package
will simply download each municipality again the first time you use it. If
you'd rather not wait for that, copy the old folder's contents over yourself.

## Not using the cache at all

With `cache = FALSE`, the data goes to a temporary file that's deleted as soon
as the call ends. Nothing is written to the cache, and every call downloads
the data again, so this only makes sense for a one-off call or when you
can't write to disk.

```{r}
cnefe <- read_cnefe(2919207, cache = FALSE)
```

## Clearing the cache

`clear_cache_muni()` removes cached municipalities, all of them or just one,
and `clear_cache_tracts()` does the same for the census tract files, all of
them or just one state:

```{r}
clear_cache_muni()           # every cached municipality
clear_cache_muni(2919207)    # only Lauro de Freitas-BA

clear_cache_tracts()         # every census tract file
clear_cache_tracts("BA")     # only the file for Bahia
```

Both functions look for the cache in the same order as everything else. If
you set `CNEFETOOLS_CACHE_DIR`, they find it on their own. If you used
`cache_dir` in a call, pass the same `cache_dir` to the cleaner, otherwise it
looks in the default folder and finds nothing to delete:

```{r}
clear_cache_muni(cache_dir = "E:/cnefe_cache")
```

If you used a version of `cnefetools` older than 0.3.0, you may also have ZIP
files at the top of the default folder, since that's how the cache used to be
stored. `clear_cache_muni()` removes those too.

## Keeping a permanent copy with `cnefe_export()`

The cache belongs to the package and can be cleared at any time. That's fine
for exploring, but not for a project that has to give the same result a year
from now. For that, `cnefe_export()` writes a municipality to a folder you
choose, and the file stays there until you delete it:

```{r}
path <- cnefe_export(2919207, path = "data/cnefe")
path
#> [1] "data/cnefe/cnefe_2022_2919207.parquet"
```

The default format is Parquet, which is smaller and faster to read than
the CSV that IBGE publishes. You can also ask for `format = "csv"` or
`format = "csv.gz"`. If the file already exists, `cnefe_export()` stops with
an error instead of overwriting it, and you can pass `overwrite = TRUE` when
you really want to replace it.

To read the copy back, give its path to `read_cnefe()` through `file`. Nothing
is downloaded, so this works offline and doesn't depend on the IBGE server:

```{r}
cnefe <- read_cnefe(file = "data/cnefe/cnefe_2022_2919207.parquet")

# Or as an sf object
cnefe_sf <- read_cnefe(file = "data/cnefe/cnefe_2022_2919207.parquet", output = "sf")
```

`file` also accepts the ZIP exactly as IBGE publishes it, as well as `.csv`
and `.csv.gz` files, so you can read a file you got some other way too.

## What reads an exported copy

Only `read_cnefe()` reads an exported file. `cnefe_counts()`, `compute_lumi()`,
`tracts_to_h3()` and `tracts_to_polygon()` always get their data through the
cache. If you need those functions to keep working without the IBGE server,
keep the cache instead of clearing it, and point `CNEFETOOLS_CACHE_DIR` at a
folder you back up along with the rest of the project.
